---
title: "Prepare table - ENVISION"
author: "[Danilo Di Leo](https://github.com/danilodileo)"
date: today
format:
  html:
    toc: true
    toc-title: "Table of Contents"
    toc-depth: 2
    toc-location: left
    number-sections: true
    code-fold: true
    code-summary: "Show Code"
    fig-width: 6
    fig-height: 4
    highlight-style: github
    self-contained: true  
editor: visual
bibliography:
  - grateful-refs.bib
---

```{r libraries}
#| label: libraries
#| message: false
#| cache: false
#| include: false
#| warning: false

library(readr)
library(dplyr)
library(tidyr)
library(lubridate)
library(purrr)
library(stringr)
library(ggplot2)
library(ggrepel)
library(kfigr)
library(knitr)
library(DT)
library(grateful)
library(patchwork)
library(broom)
library(gt)
library(tibble)
library(corrr)
library(forcats)
library(shiny)
library(grid)
library(ggvenn)
library(vegan)
library(reshape2)
library(ggsignif)
library(readxl)
library(ggcorrplot)
library(FactoMineR)
library(factoextra)
library(missMDA)
library(kableExtra) # For tables
library(compositions) # For CLR transformation
library(pheatmap)
library(gridExtra)
library(tidytext)  # per reorder_within() e scale_x_reordered()
library(shadowtext)
library(reactable)
```

```{r constant}
# Set COG categories long names
cog_mapping <- c(
   'A' = 'RNA_process',
   'B' = 'Chromatin_Struc',
   'C' = 'Energy_prod',
   'D' = 'Cell_cycle',
   'E' = 'AA_metab',
   'F' = 'Nucl_metab',
   'G' = 'Carbohy_metab',
   'H' = 'Coenz_metab',
   'I' = 'Lipid_metab',
   'J' = 'Tranlsation',
   'K' = 'Transcription',
   'L' = 'Replication',
   'M' = 'Cell_wall',
   'N' = 'Cell_motility',
   'O' = 'Post-transl_mod',
   'P' = 'Inorg_ion_transp',
   'Q' = 'Sec_Structure',
  'T' = 'Signal_Transduc',
  'U' = 'Intracel_traf',
  'Y' = 'Nucl_structure',
  'Z' = 'Cytoskeleton',
  'R' = 'Gen_Func',
  'S' = 'Func_unknown',
  '-' = 'Func_unknown'
 )

# sample order
sample_order <- c(
  "ENV1_D1_S3", "ENV1_D3_S3", "ENV1_D5_S3", "ENV1_D7_S3",
  "ENV1_D1_S6", "ENV1_D3_S6", "ENV1_D5_S6", "ENV1_D7_S6",
  "ENV2_D1_S3", "ENV2_D3_S3", "ENV2_D5_S3", "ENV2_D7_S3",
  "ENV2_D1_S6", "ENV2_D3_S6", "ENV2_D5_S6", "ENV2_D7_S6",
  "ENV3_D1_S3", "ENV3_D3_S3", "ENV3_D5_S3", "ENV3_D7_S3",
  "ENV3_D1_S6", "ENV3_D3_S6", "ENV3_D5_S6", "ENV3_D7_S6"
)
```

## Code explanation

This quarto document produces the intermediate summary tables consumed by the
other analysis scripts in this repository
(`Overall_community_composition_and_environmental_variable_analysis.qmd`,
`fine-scale-analysis.qmd`, `untargeted_analysis.qmd`).

The tables are produced from raw tables output of:

-   nf-core/magmap

-   eggnog-mapper

-   targeted hmmer profiles

## Formatting Count table

Sample names encode site and season (e.g. `ENV2_D3_S6`: ENVISION cruise 2 =
Spring, day 3, station S6 = Offshore); a fourth, non-cruise batch of triplicate
mesocosm incubations is tagged `_M0`-`_M3` and is averaged down to a single
`_M` value per ORF below before the two groups are combined.

```{r count-table}
#| message: false
#| cache: false
#| include: false
#| warning: false
# Read counts table and format them
counts <- read_tsv(
  'summary_tables/magmap.CDS.counts.tsv.gz',
  col_types = cols(
    .default = col_double(), orf = col_character(), chr = col_character(),
   sample = col_character(), strand = col_character()
    )) %>%
  select( -start, -end, -chr, -strand)

# I need to get the average for the mesocosm samples
# Step 1: Process `M` triplicates
triplicates <- counts %>%
  filter(str_detect(sample, "_M[0-3]$")) %>%
  mutate(sample = str_remove(sample, "_M[0-3]$") %>% paste0("_M")) %>%
  group_by(orf, length, sample) %>%
  summarize(
    count   = mean(count),
    tpm     = mean(tpm),
    sdcount = sd(count),
    sdtpm   = sd(tpm), .groups = 'drop')


# Step 2: Combine with the original dataframe
counts <- counts %>%
  filter(!str_detect(sample, "_M[0-3]$")) %>%  # Keep non-triplicate samples
  mutate( sdcount = 0,
          sdtpm   = 0) %>%
  arrange(count, sdcount, tpm, sdtpm) %>%
  bind_rows(triplicates)

rm(triplicates)

# bind counts with genomes_index + formatting the table
counts <- counts %>%
  mutate(
    season = case_when(
    startsWith(sample, "ENV1") ~ "Winter",
    startsWith(sample, "ENV2") ~ "Spring",
    startsWith(sample, "ENV3") ~ "Summer",
    TRUE ~ "Unknown"  # Optional: label for cases that don't match any pattern
    ),
    season = factor(season, levels = c("Winter", "Spring", "Summer")),
    site = case_when(
      str_detect(sample, "S3") ~ "Coast",      # When "S3" is present
      str_detect(sample, "S6") ~ "Offshore",   # When "S6" is present
      str_detect(sample, "M") ~ "Mesocosm",   # When "M" is present
      TRUE ~ "Other"                          # Optional: for any other case
    ),
    site = factor(site, levels = c("Offshore", "Coast", "Mesocosm"))
  ) %>%
  left_join(
    read_tsv(
      'summary_tables/magmap.genomes2orfs.tsv.gz', # load genome index for taxonomy annotation
      show_col_types = FALSE
    ),
    by = "orf"
    )

orfs <- counts %>%
  distinct(orf) %>%
  pull(orf)
```

## Add taxonomy

```{r add-taxonomy-to-count-table}
#| message: false
#| cache: false
#| include: false
#| warning: false

# GTDB reports one semicolon-delimited lineage string per genome
# (e.g. "d__Bacteria;p__...;f__Pelagibacteraceae;..."); this splits it into
# one ranked column per taxon and strips the "x__" rank prefix GTDB adds.
parse_gtdb_taxonomy <- function(df) {
  df %>%
    separate(
      gtdb_taxonomy,
      into = c("domain", "phylum", "class", "order", "family", "genus", "species"),
      sep = ";",
      fill = "right"
    ) %>%
    mutate(across(
      everything(),
      ~ str_remove(., "^[a-z]__") %>% na_if("") %>% replace_na("unclassified")
    ))
}

# NCBI/GTDB reference genomes (magmap's own taxonomy assignment)
ncbi.taxonomy <- read_tsv('summary_tables/magmap.genome_metadata.tsv.gz',
         show_col_types = FALSE
         ) %>%
  select(accno, gtdb_taxonomy) %>%
  parse_gtdb_taxonomy() %>%
  filter(!str_detect(accno, 'metaba*'),
         !str_detect(accno, 'conco*'))

# Locally assembled/binned genomes, classified separately with GTDB-Tk
# (archaea and bacteria are reported in separate GTDB-Tk output files)
arc.taxonomy <- read_tsv('gtdbtk.ar53.summary.tsv', show_col_types = FALSE) %>%
  select(user_genome, classification) %>%
  rename( accno = user_genome,
          gtdb_taxonomy = classification) %>%
  parse_gtdb_taxonomy() %>%
  mutate(accno = str_remove(accno, ".fa"))

bac.taxonomy <- read_tsv('gtdbtk.bac120.summary.tsv', show_col_types = FALSE) %>%
  select(user_genome, classification) %>%
  rename( accno = user_genome,
          gtdb_taxonomy = classification) %>%
  parse_gtdb_taxonomy() %>%
  mutate(accno = str_remove(accno, ".fa"))

# Bind taxonomy together
taxonomy <- bind_rows(ncbi.taxonomy,arc.taxonomy,bac.taxonomy)

# write taxonomy table
write_tsv(taxonomy,'summary_tables/taxonomy.tsv.gz')
```

## Merge counts and taxonomy together

```{r merge-counts-tax}
#| message: false
#| cache: false
#| include: false
#| warning: false

counts.tax <- counts %>%
  left_join(taxonomy, join_by(accno),
            relationship = "many-to-many")
```

## Eggnog-mapper functional table

```{r function-table}
#| message: false
#| cache: false
#| include: false
#| warning: false
# Add function and remove the previous annotation

# eggnog-mapper annotation TSVs have either 21 or 24 columns depending on
# whether the ORF matched a broader ("best_og_*") annotation level; group
# files by column count so each group can be read with consistent names.
count_columns <- function(file) {
  read_tsv(file, comment = "##", show_col_types = FALSE) %>% ncol()
}

files_by_column_count <- function(files, n) {
  tibble(file = files, num_cols = map_int(files, count_columns)) %>%
    filter(num_cols == n) %>%
    pull(file)
}

# spot which files have different number of columns
files <- Sys.glob('summary_tables/*/*annotations.gz')
file.with.21.columns <- files_by_column_count(files, 21)
file.with.24.columns <- files_by_column_count(files, 24)

eggnog <-  rbind(
  file.with.24.columns %>% # load all annotated ORFs
    read_tsv(id = 'fname', comment = "##", show_col_types = FALSE) %>%
    rename_with(~ tolower(sub("^#", "", .)))  %>%
    rename( orf = 'query',
          max_annot_lvl = 'best_og_name',
          cog_category = 'best_og_cat',
          description = 'best_og_desc') %>%
    mutate( project = 'gtdb') %>%
    select(-fname, -narr_og_name, -narr_og_cat, -narr_og_desc),
  file.with.21.columns %>%
    read_tsv(id = 'fname', comment = "##", show_col_types = FALSE) %>%
    rename_with(~ tolower(sub("^#", "", .)))  %>%
    rename( orf = 'query') %>%
    mutate( project = 'gtdb') %>%
    select(-fname)) %>%
  filter(orf %in% orfs)

# write_tsv(eggnog.annotation, "summary_tables/eggnog_annotation.tsv.gz")
```

```{r leftover-annotation}

# ORFs that eggnog-mapper's first pass missed, annotated in a follow-up run
# (same column-count grouping as function-table above)
files <- Sys.glob('summary_tables/missing-emapper-annotation/*annotations.gz')
file.with.21.columns <- files_by_column_count(files, 21)
file.with.24.columns <- files_by_column_count(files, 24)

eggnog.annotation <-  file.with.21.columns %>%
    read_tsv(id = 'fname', comment = "##", show_col_types = FALSE) %>%
    rename_with(~ tolower(sub("^#", "", .)))  %>%
    rename( orf = 'query') %>%
    mutate( project = 'gtdb') %>%
    select(-fname) %>%
  filter(orf %in% orfs)

#write_tsv(eggnog.total, "summary_tables/eggnog_annotation_total.tsv.gz")
```

## KOfamscan functional table

```{r load-kofam-tables}
files <- Sys.glob('summary_tables/kofamscan/*')

all_data <- files %>%
  map_dfr(~ read_tsv(.x, show_col_types = FALSE), .id = "fname") %>%
  filter(sign == '*') %>%
  rename(orf = 'accno') %>%
  select(-fname, -thrshld, -score, -evalue)

write_tsv(all_data, 'kofam_annotation_top30genomes.tsv.gz')
```

## Targeted hmmer profiles table

```{r target-annotation-table}
#| message: false
#| cache: false
#| include: false
#| warning: false
# load rhodopsin and other genes from José analysis
annotation <- read_delim("targeted-genes.csv",delim = ";", col_names = TRUE,
                    show_col_types = FALSE) %>%
  select(-'No. peptides') %>%
  pivot_longer(cols = -genome, names_to = "annotation", values_to = "orf") %>%
  filter(orf != 0) %>%
  separate_rows(orf, sep = ",\\s*") %>%
   mutate(annotation = case_when(
    str_detect(annotation, "rhodopsin") ~ "rhodopsin",
    str_detect(annotation, "PEP") ~ "pepcase",
    str_detect(annotation, "RecA") ~ "reca",
    str_detect(annotation, "PufM") ~ "pufm",
    str_detect(annotation, "pyr ca") ~ "pyrcarb",
    str_detect(annotation, "crotonyl-CoA") ~ "ccr",
    str_detect(annotation, "isocitrat") ~ "isoly",
    str_detect(annotation, "TonB") ~ "tonb",
    TRUE ~ annotation ) ) %>%
  mutate(genome = genome %>%
           # First lowercase initial M or C
           str_replace("^M", "m") %>%
           str_replace("^C", "c") %>%
           # Handle GCA/GCF cases - replace only the FIRST underscore after them
           str_replace("^(GCA|GCF)_(.*?)_", "\\1_\\2.") %>%
           # Replace any remaining underscores with dots
           str_replace_all("_", ".") %>%
           # Special case: if F or A precedes a dot, make it underscore
           str_replace_all("([FA])\\.", "\\1_"))

# write_tsv(annot, 'summary_tables/targeted_genes.tsv.gz')
```

## Load metadata

```{r load-metadata}
#| message: false
#| cache: false
#| include: false
#| warning: false
# Load metadata (Coast and Offshore sites)
metadata <- read_excel("data/metatada_summarized.xlsx") %>%
  select(-bb, -po4, -nh4, -sio2, -tin, -no3, -chla_less_3um, -chla_min_3um, -chla_total, -ab, -no2)

superficial.metadata <- read_excel('data/all_Superficial_Metadata_ENVISION.xlsx') %>%
  select(-cruise, -cruise_station, -Season_Station, -Condition, -day, -station, -Depth, -Bacterial_biomass)

envision.metadata <- superficial.metadata %>%
  left_join(metadata, join_by(sample)) %>%
  rename(temperature = "Temperature", tin = 'TIN',
         chla_tot = "Chlorophyll_a",
         no3 = "NO3", po4 = "PO4",
         pa = "Bacterial_abundance", sio2 = 'SiO2',
         nh4 = 'NH4', no2 = "NO2",
         chla_less_3um = 'Chla_0.2', chla_min_3um = 'Chla_3')

write_tsv(envision.metadata, './data/envision_metadata.tsv.gz')
```

## Create complete count table

```{r create-complete-count-table}
# Add day and site/season filters on top of the counts+taxonomy table
# built above (counts.tax, from the "Merge counts and taxonomy" section)
counts <- counts.tax %>%
  mutate(
    day = case_when(
      str_detect(sample, "D0") ~ "0",
      str_detect(sample, "D1") ~ "1",
      str_detect(sample, "D3") ~ "3",
      str_detect(sample, "D5") ~ "5",
      str_detect(sample, "D7") ~ "7",
      TRUE ~ "Other"                          # Optional: for any other case
    )
  )

# Remove mesocosm experiment
counts <- counts %>%
  filter( site != 'Mesocosm') %>%
  filter( domain != 'NA')


# create one whole table
eggnog.total <- bind_rows(eggnog, eggnog.annotation)

# Join targeted annotation table to counts and taxonomy
counts <- counts %>% left_join(eggnog.total, join_by(orf))

# Load targeted annotation table
annotation <- read_tsv('summary_tables/targeted_genes.tsv.gz', show_col_types = FALSE)

# Join targeted annotation table to counts and taxonomy
counts <- counts %>%
  left_join(annotation %>%
              select(
                -genome), join_by(orf))

# create table with only genomes expressing rhodopsin
counts.rhodo.genomes <- counts %>%
 filter(annotation == "rhodopsin") %>%
 distinct(genome) %>%
 inner_join(counts, join_by(genome))

# write_tsv(counts.rhodo.genomes, 'summary_tables/counts-rhodo-genomes-taxonomy-function.tsv.gz')

# write_tsv(counts, 'summary_tables/counts-taxonomy-function.tsv.gz')
```
