---
title: "Prepare table - ENVISION"
author: "[Danilo Di Leo](https://github.com/danilodileo)"
date: today
format:
  html:
    toc: true
    toc-title: "Table of Contents"
    toc-depth: 2
    toc-location: left
    number-sections: true
    code-fold: true
    code-summary: "Show Code"
    fig-width: 6
    fig-height: 4
    highlight-style: github
    self-contained: true  
editor: visual
bibliography:
  - grateful-refs.bib
---

```{r libraries}
#| label: libraries
#| message: false
#| cache: false
#| include: false
#| warning: false

library(readr)
library(dplyr)
library(tidyr)
library(lubridate)
library(purrr)
library(stringr)
library(ggplot2)
library(ggrepel)
library(kfigr)
library(knitr)
library(DT)
library(grateful)
library(patchwork)
library(broom)
library(gt)
library(tibble)
library(corrr)
library(forcats)
library(shiny)
library(grid)
library(ggvenn)
library(vegan)
library(reshape2)
library(ggsignif)
library(readxl)
library(ggcorrplot)
library(FactoMineR)
library(factoextra)
library(missMDA)
library(kableExtra) # For tables
library(compositions) # For CLR transformation
library(pheatmap)
library(gridExtra)
library(tidytext)  # per reorder_within() e scale_x_reordered()
library(shadowtext)
library(gt)
library(reactable)
```

```{r constant }
# Set COG categories long names
cog_mapping <- c(
   'A' = 'RNA_process',
   'B' = 'Chromatin_Struc',
   'C' = 'Energy_prod',
   'D' = 'Cell_cycle',
   'E' = 'AA_metab',
   'F' = 'Nucl_metab',
   'G' = 'Carbohy_metab',
   'H' = 'Coenz_metab',
   'I' = 'Lipid_metab',
   'J' = 'Tranlsation',
   'K' = 'Transcription',
   'L' = 'Replication',
   'M' = 'Cell_wall',
   'N' = 'Cell_motility',
   'O' = 'Post-transl_mod',
   'P' = 'Inorg_ion_transp',
   'Q' = 'Sec_Structure',
  'T' = 'Signal_Transduc',
  'U' = 'Intracel_traf',
  'Y' = 'Nucl_structure',
  'Z' = 'Cytoskeleton',
  'R' = 'Gen_Func',
  'S' = 'Func_unknown',
  '-' = 'Func_unknown'
 )

# sample order
sample_order <- c(
  "ENV1_D1_S3", "ENV1_D3_S3", "ENV1_D5_S3", "ENV1_D7_S3",
  "ENV1_D1_S6", "ENV1_D3_S6", "ENV1_D5_S6", "ENV1_D7_S6",
  "ENV2_D1_S3", "ENV2_D3_S3", "ENV2_D5_S3", "ENV2_D7_S3",
  "ENV2_D1_S6", "ENV2_D3_S6", "ENV2_D5_S6", "ENV2_D7_S6",
  "ENV3_D1_S3", "ENV3_D3_S3", "ENV3_D5_S3", "ENV3_D7_S3",
  "ENV3_D1_S6", "ENV3_D3_S6", "ENV3_D5_S6", "ENV3_D7_S6"
)
```

## Code explanation

This quarto document is used in order to produce tables that can be used in the main analysis in other quarto documents (e.g. 'Fluctuation in photoheterotrophic community.qmd')

The tables are produced from raw tables output of:

-   nf-core/magmap

-   eggnog-mapper

-   targeted hmmer profiles

## Formatting Count table

```{r count-table}
#| message: false
#| cache: false
#| include: false
#| warning: false
# Read counts table and format them
counts <- read_tsv(
  'summary_tables/magmap.CDS.counts.tsv.gz',
  col_types = cols(
    .default = col_double(), orf = col_character(), chr = col_character(),
   sample = col_character(), strand = col_character()
    )) %>%
  select( -start, -end, -chr, -strand)

# I need to get the average for the mesocosm samples
# Step 1: Process `M` triplicates
triplicates <- counts %>%
  filter(str_detect(sample, "_M[0-3]$")) %>%
  mutate(sample = str_remove(sample, "_M[0-3]$") %>% paste0("_M")) %>%
  group_by(orf, length, sample) %>%
  summarize(
    count   = mean(count),
    tpm     = mean(tpm),
    sdcount = sd(count),
    sdtpm   = sd(tpm), .groups = 'drop')


# Step 2: Combine with the original dataframe
counts <- counts %>%
  filter(!str_detect(sample, "_M[0-3]$")) %>%  # Keep non-triplicate samples
  mutate( sdcount = 0,
          sdtpm   = 0) %>%
  arrange(count, sdcount, tpm, sdtpm) %>%
  bind_rows(triplicates)

rm(triplicates)

# bind counts with genomes_index + formatting the table
counts <- counts %>%
  mutate(
    season = case_when(
    startsWith(sample, "ENV1") ~ "Winter",
    startsWith(sample, "ENV2") ~ "Spring",
    startsWith(sample, "ENV3") ~ "Summer",
    TRUE ~ "Unknown"  # Optional: label for cases that don't match any pattern
    ),
    season = factor(season, levels = c("Winter", "Spring", "Summer")),
    site = case_when(
      str_detect(sample, "S3") ~ "Coast",      # When "S3" is present
      str_detect(sample, "S6") ~ "Offshore",   # When "S6" is present
      str_detect(sample, "M") ~ "Mesocosm",   # When "M" is present
      TRUE ~ "Other"                          # Optional: for any other case
    ),
    site = factor(site, levels = c("Offshore", "Coast", "Mesocosm"))
  ) %>%
  left_join(
    read_tsv(
      'summary_tables/magmap.genomes2orfs.tsv.gz', # load genome index for taxonomy annotation
      show_col_types = FALSE
    ),
    by = "orf"
    )

orfs <- counts %>%
  distinct(orf) %>%
  pull(orf)
```

## Add taxonomy

```{r add-taxonomy-to-count-table}
#| message: false
#| cache: false
#| include: false
#| warning: false
# Read and format ncbi taxonomy table
ncbi.taxonomy <- read_tsv('summary_tables/magmap.genome_metadata.tsv.gz',
         show_col_types = FALSE
         ) %>%
  select(accno, gtdb_taxonomy) %>%
  separate(
    gtdb_taxonomy,
    into = c("domain", "phylum", "class", "order", "family", "genus", "species"),
    sep = ";",
    fill = "right"
  ) %>%
  mutate(across(
    everything(),
    ~ str_remove(., "^[a-z]__") %>% na_if("") %>% replace_na("unclassified")
  )) %>%
  filter(!str_detect(accno, 'metaba*'),
         !str_detect(accno, 'conco*'))

# get local archaea taxonomy
arc.taxonomy <- read_tsv('gtdbtk.ar53.summary.tsv', show_col_types = FALSE) %>%
  select(user_genome, classification) %>%
  rename( accno = user_genome,
          gtdb_taxonomy = classification) %>%
  separate(
    gtdb_taxonomy,
    into = c("domain", "phylum", "class", "order", "family", "genus", "species"),
    sep = ";",
    fill = "right"
  ) %>%
  mutate(across(
    everything(),
    ~ str_remove(., "^[a-z]__") %>% na_if("") %>% replace_na("unclassified")
  )) %>%
  mutate(accno = str_remove(accno, ".fa"))

# Get local bacteria taxonomy
bac.taxonomy <- read_tsv('gtdbtk.bac120.summary.tsv', show_col_types = FALSE) %>%
  select(user_genome, classification) %>%
  rename( accno = user_genome,
          gtdb_taxonomy = classification) %>%
  separate(
    gtdb_taxonomy,
    into = c("domain", "phylum", "class", "order", "family", "genus", "species"),
    sep = ";",
    fill = "right"
  ) %>%
  mutate(across(
    everything(),
    ~ str_remove(., "^[a-z]__") %>% na_if("") %>% replace_na("unclassified")
  )) %>%
  mutate(accno = str_remove(accno, ".fa"))

# Bind taxonomy together
taxonomy <- bind_rows(ncbi.taxonomy,arc.taxonomy,bac.taxonomy)

# write taxonomy table
write_tsv(taxonomy,'summary_tables/taxonomy.tsv.gz')
```

## Merge counts and taxonomy together

```{r merge-counts-tax }
#| message: false
#| cache: false
#| include: false
#| warning: false

counts.tax <- counts %>%
  left_join(taxonomy, join_by(accno),
            relationship = "many-to-many")

#write_tsv(counts.tax,'counts-taxonomy.tsv.gz')
```

## Eggnog-mapper functional table

```{r function-table}
#| message: false
#| cache: false
#| include: false
#| warning: false
# Add function and remove the previous annotation

# spot which files have different number of columns
files <- Sys.glob('summary_tables/*/*annotations.gz')

# # Function to count columns in each file
count_columns <- function(file) {
  num_cols <- read_tsv(file, comment = "##", show_col_types = FALSE) %>%
  ncol()
  tibble(file = file, num_cols = num_cols)
}

# # Apply the function to each file and collect the results
file_column_counts <- map_dfr(files, count_columns)
 
# Pull files with 21 columns
file.with.21.columns <- file_column_counts %>%
filter( num_cols == 21) %>%
select(-num_cols) %>%
pull(file)

 
# Pull files with 24 columns
file.with.24.columns <- file_column_counts %>%
 filter( num_cols == 24) %>%
 select(-num_cols) %>%
 pull(file)

eggnog <-  rbind(
  file.with.24.columns %>% # load all annotated ORFs
    read_tsv(id = 'fname', comment = "##", show_col_types = FALSE) %>%
    rename_with(~ tolower(sub("^#", "", .)))  %>%
    rename( orf = 'query',
          max_annot_lvl = 'best_og_name',
          cog_category = 'best_og_cat',
          description = 'best_og_desc') %>%
    mutate( project = 'gtdb') %>%
    select(-fname, -narr_og_name, -narr_og_cat, -narr_og_desc),
  file.with.21.columns %>%
    read_tsv(id = 'fname', comment = "##", show_col_types = FALSE) %>%
    rename_with(~ tolower(sub("^#", "", .)))  %>%
    rename( orf = 'query') %>%
    mutate( project = 'gtdb') %>%
    select(-fname)) %>%
  filter(orf %in% orfs)

# write_tsv(eggnog.annotation, "summary_tables/eggnog_annotation.tsv.gz")
```

```{r leftover-annotation}

# spot which files have different number of columns
files <- Sys.glob('summary_tables/missing-emapper-annotation/*annotations.gz')

# # Function to count columns in each file
count_columns <- function(file) {
  num_cols <- read_tsv(file, comment = "##", show_col_types = FALSE) %>%
  ncol()
  tibble(file = file, num_cols = num_cols)
}

# # Apply the function to each file and collect the results
file_column_counts <- map_dfr(files, count_columns)
 
# Pull files with 21 columns
file.with.21.columns <- file_column_counts %>%
filter( num_cols == 21) %>%
select(-num_cols) %>%
pull(file)

 
# Pull files with 24 columns
file.with.24.columns <- file_column_counts %>%
 filter( num_cols == 24) %>%
 select(-num_cols) %>%
 pull(file)

eggnog.annotation <-  file.with.21.columns %>%
    read_tsv(id = 'fname', comment = "##", show_col_types = FALSE) %>%
    rename_with(~ tolower(sub("^#", "", .)))  %>%
    rename( orf = 'query') %>%
    mutate( project = 'gtdb') %>%
    select(-fname) %>%
  filter(orf %in% orfs)

#write_tsv(eggnog.total, "summary_tables/eggnog_annotation_total.tsv.gz")
```

## KOfamscan functional table

```{r load-kofam-tables}
files <- Sys.glob('summary_tables/kofamscan/*')

all_data <- files %>%
  map_dfr(~ read_tsv(.x, show_col_types = FALSE), .id = "fname") %>%
  filter(sign == '*') %>%
  rename(orf = 'accno') %>%
  select(-fname, -thrshld, -score, -evalue)

write_tsv(all_data, 'kofam_annotation_top30genomes.tsv.gz')
```

## Targeted hmmer profiles table

```{r target-annotation-table }
#| message: false
#| cache: false
#| include: false
#| warning: false
# load rhodopsin and other genes from José analysis
annotation <- read_delim("targeted-genes.csv",delim = ";", col_names = TRUE,
                    show_col_types = FALSE) %>%
  select(-'No. peptides') %>%
  pivot_longer(cols = -genome, names_to = "annotation", values_to = "orf") %>%
  filter(orf != 0) %>%
  separate_rows(orf, sep = ",\\s*") %>%
   mutate(annotation = case_when(
    str_detect(annotation, "rhodopsin") ~ "rhodopsin",
    str_detect(annotation, "PEP") ~ "pepcase",
    str_detect(annotation, "RecA") ~ "reca",
    str_detect(annotation, "PufM") ~ "pufm",
    str_detect(annotation, "pyr ca") ~ "pyrcarb",
    str_detect(annotation, "crotonyl-CoA") ~ "ccr",
    str_detect(annotation, "isocitrat") ~ "isoly",
    str_detect(annotation, "TonB") ~ "tonb",
    TRUE ~ annotation ) ) %>%
  mutate(genome = genome %>%
           # First lowercase initial M or C
           str_replace("^M", "m") %>%
           str_replace("^C", "c") %>%
           # Handle GCA/GCF cases - replace only the FIRST underscore after them
           str_replace("^(GCA|GCF)_(.*?)_", "\\1_\\2.") %>%
           # Replace any remaining underscores with dots
           str_replace_all("_", ".") %>%
           # Special case: if F or A precedes a dot, make it underscore
           str_replace_all("([FA])\\.", "\\1_"))

# write_tsv(annot, 'summary_tables/targeted_genes.tsv.gz')
```

## Load metadata

```{r load-metadata}
#| message: false
#| cache: false
#| include: false
#| warning: false
# Load meatadata Coast and Offshore
metadata <- read_excel("data/metatada_summarized.xlsx") %>%
  select(-bb, -po4, -nh4, -sio2, -tin, -no3, -chla_less_3um, -chla_min_3um, -chla_total, -ab, -no2)

erick.metadata <- read_excel('data/all_Superficial_Metadata_ENVISION.xlsx') %>%
  select(-cruise, -cruise_station, -Season_Station, -Condition, -day, -station, -Depth, -Bacterial_biomass)

more.metadata <- erick.metadata %>%
  left_join(metadata, join_by(sample)) %>%
  rename(temperature = "Temperature", tin = 'TIN',
         chla_tot = "Chlorophyll_a",
         no3 = "NO3", po4 = "PO4",
         pa = "Bacterial_abundance", sio2 = 'SiO2',
         nh4 = 'NH4', no2 = "NO2",
         chla_less_3um = 'Chla_0.2', chla_min_3um = 'Chla_3')

write_tsv(more.metadata, './data/envision_metadata.tsv.gz')
```

## Create complete count table

```{r create-complete-count-table}
# Read count table
counts <- read_tsv('counts-taxonomy.tsv.gz', show_col_types = FALSE) %>%
  mutate(
    day = case_when(
      str_detect(sample, "D0") ~ "0",
      str_detect(sample, "D1") ~ "1",
      str_detect(sample, "D3") ~ "3",
      str_detect(sample, "D5") ~ "5",
      str_detect(sample, "D7") ~ "7",
      TRUE ~ "Other"                          # Optional: for any other case
    )
  )

# Remove mesocosm experiment
counts <- counts %>%
  filter( site != 'Mesocosm') %>%
  filter( domain != 'NA')


# create one whole table
eggnog.total <- bind_rows(eggnog, filtered.eggnog.annotation)

# Join targeted annotation table to counts and taxonomy
counts <- counts %>% left_join(eggnog.total, join_by(orf))

# Load targeted annotation table
annotation <- read_tsv('summary_tables/targeted_genes.tsv.gz', show_col_types = FALSE)

# Join targeted annotation table to counts and taxonomy
counts <- counts %>%
  left_join(annotation %>%
              select(
                -genome), join_by(orf))

# create table with only genomes expressing rhodopsin
counts.rhodo.genomes <- counts %>%
 filter(annotation == "rhodopsin") %>%
 distinct(genome) %>%
 inner_join(counts, join_by(genome))

# write_tsv(counts.rhodo.genomes, 'summary_tables/counts-rhodo-genomes-taxonomy-function.tsv.gz')

# write_tsv(counts, 'summary_tables/counts-taxonomy-function.tsv.gz')
```
